---
title: "Figure 6. Cell-cell communication networks in Sjögren's syndrome PBMCs"
author: "Cheng"
date: "2026/04/21"
output: 
  html_document:
    toc: false
    toc_float: false
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
    fig_caption: true
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 1. INITIALIZATION AND SETUP
# ============================================================================
knitr::opts_chunk$set(
  echo = TRUE,           
  warning = TRUE,        
  message = TRUE,       
  error = TRUE,          
  fig.width = 10,        
  fig.height = 7,        
  fig.align = "center",  
  cache = FALSE,         
  dpi = 300              
)

# 清理工作空间
rm(list = ls())
gc()

# 设置输出目录
output_dir <- "."
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 设置随机种子保证可重复性
set.seed(123)
```

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 2. LOAD REQUIRED LIBRARIES
# ============================================================================
#CellChat 相关
library(CellChat)
library(NMF)
library(ggalluvial)
library(patchwork)

#数据处理
library(tidyverse)
library(Seurat)
library(SeuratData)
library(here)

#可视化
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(ComplexHeatmap)
library(circlize)

#检查并安装缺失包
required_packages <- c("CellChat", "NMF", "ggalluvial", "tidyverse", "Seurat")
missing_packages <- required_packages[!required_packages %in% installed.packages()]
if (length(missing_packages) > 0) {
message("Installing missing packages: ", paste(missing_packages, collapse = ", "))
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install(missing_packages, update = FALSE)
}
```

<br>

# **Figure 6: Cell-cell communication networks in Sjögren's syndrome**

**Figure Legend:** Comprehensive cell-cell communication analysis of PBMCs from healthy controls (HC) and Sjögren's syndrome (SjS) patients. (a) Overview of inferred ligand-receptor interactions. (b) Signaling pathway enrichment analysis. (c) Differential cell-cell communication between HC and SjS. (d) Key signaling pathways contributing to disease progression.

---

## **a. Overview of ligand-receptor interactions**

**(a) Global view of ligand-receptor interactions in HC and SjS PBMCs.** Network diagrams showing the overall communication strength between different immune cell types. The thickness of edges represents interaction strength, and node size indicates the number of cells.

```{r fig1a, fig.width=9, fig.height=8, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 6A: CELL TYPE COUNT BARPLOT (HC vs SjS)
# ============================================================================

# 假设 scRNA_harmony 已加载，且包含 Group (HC/SjS) 和 Minor.CellType 列
library(ggplot2)
library(dplyr)

# 提取每个组的小细胞类型计数
cell_counts <- scRNA_harmony@meta.data %>%
  group_by(Group, Minor.CellType) %>%
  summarise(n = n(), .groups = 'drop') %>%
  mutate(Minor.CellType = factor(Minor.CellType, levels = unique(Minor.CellType)))  # 保持原始顺序

# 绘图
p6a <- ggplot(cell_counts, aes(x = Minor.CellType, y = n, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("HC" = "#1F77B4", "SjS" = "#D62728")) +
  labs(
    title = "Cell type distribution in HC vs SjS",
    subtitle = "Count per minor cell type",
    x = "Minor Cell Type",
    y = "Number of Cells"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "top"
  ) +
  guides(fill = guide_legend(title = "Group", nrow = 1))

# 保存
ggsave(
  filename = here(output_dir, "Figure6A_cell_counts.pdf"),
  plot = p2a,
  width = 8,
  height = 6,
  device = "pdf"
)
```

---

## **b. Signaling pathway heatmap**
**(b) Heatmap of signaling pathway communication weights.

```{r fig1b, fig.width=9, fig.height=8, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 6B: SIGNALING PATHWAY HEATMAP (WEIGHT)
# ============================================================================

# 从 cellchat@netP 中提取权重矩阵
library(NMF)
library(reshape2)

# 提取 netP 权重矩阵（signaling pathway level）
netP_weights <- cellchat@netP$weight
rownames(netP_weights) <- cellchat@netP$pathway
colnames(netP_weights) <- paste0(cellchat@netP$sender, "→", cellchat@netP$receiver)

# 对行（pathway）和列（pair）进行排序（可选：按总分排序）
pathway_order <- rownames(netP_weights)[order(rowSums(netP_weights), decreasing = TRUE)]
pair_order <- colnames(netP_weights)[order(colSums(netP_weights), decreasing = TRUE)]

netP_weights_sorted <- netP_weights[pathway_order, pair_order]

# 转换为长格式用于绘图
plot_data <- melt(as.matrix(netP_weights_sorted)) %>%
  rename(Pathway = Var1, Pair = Var2, Weight = value)

# 绘图（使用 NMF 的 aheatmap 或 ggplot2）
# 这里使用 ggplot2 实现热图
p2b <- ggplot(plot_data, aes(x = Pair, y = Pathway, fill = Weight)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colors = rev(brewer.pal(11, "RdBu")),
    limits = c(0, max(plot_data$Weight)),
    na.value = "white",
    name = "Weight"
  ) +
  labs(
    title = "Signaling pathway communication weights in HC",
    x = "Sender → Receiver",
    y = "Signaling Pathway"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "right"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))

# 保存
ggsave(
  filename = here(output_dir, "Figure6B_heatmap_weight.pdf"),
  plot = p2b,
  width = 10,
  height = 8,
  device = "pdf"
)
```

---

