---
title: "Figure 2. Characterization of CD8⁺ T Cell Subsets in Sjögren's Syndrome"
author: "Cheng"
date: "2026/04/21"
output: 
  html_document:
    toc: false
    toc_float: false
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
    fig_caption: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 1. INITIALIZATION AND SETUP
# ============================================================================
knitr::opts_chunk$set(
  echo = TRUE,           
  warning = TRUE,        
  message = TRUE,        
  error = TRUE,          
  fig.width = 10,        
  fig.height = 7,        
  fig.align = "center",  
  cache = FALSE,        
  dpi = 300              
)

# 清理工作空间
rm(list = ls())
gc()

# 设置输出目录
output_dir <- "."
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 设置随机种子保证可重复性
set.seed(123)
```

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 2. LOAD REQUIRED LIBRARIES
# ============================================================================
# 核心分析包

library(Seurat) 
library(plotrix) 

# 可视化包
library(ggplot2) 
library(patchwork) 
library(RColorBrewer) 
library(viridis) 
library(ggrepel) 
library(ggstatsplot) 

# 功能分析包
library(clusterProfiler) 
library(org.Hs.eg.db) 
library(enrichplot) 

# 数据处理包
library(dplyr) 
library(tidyr) 
library(tibble) 

# 检查并安装缺失包
required_packages <- c("Seurat", "ggplot2", "plotrix", "dplyr", "clusterProfiler")
missing_packages <- required_packages[!required_packages %in% installed.packages()]
if (length(missing_packages) > 0) {
message("Installing missing packages: ", paste(missing_packages, collapse = ", "))
BiocManager::install(missing_packages, update = FALSE)
}
```

<br>

# **Figure 2: Characterization of CD8⁺ T Cell Subsets in Sjögren's Syndrome**

**Figure Legend:** Analysis of CD8⁺ T cell heterogeneity in Sjögren's syndrome. (a) Proportional distribution of four CD8⁺ T cell subsets. (b) Violin plots showing expression of 18 marker genes across subsets. (c) Dot plot visualizing expression patterns of 45 selected genes in cluster3. (d) Volcano plot of differentially expressed genes in cluster3 between SjS and HC. (e) Volcano plot of differentially expressed genes in cluster19 between SjS and HC. (f) GO enrichment analysis of differentially expressed genes from CD8⁺ T cell clusters.

---

## **a. Proportion of CD8⁺ T cell subset**

**(a) 3D pie chart showing the relative abundance of four CD8⁺ T cell clusters (cluster2, cluster3, cluster11, cluster19) identified in PBMCs.** Cluster3 (39%) represents the most abundant subset, followed by cluster2 (43%), cluster11 (15%), and cluster19 (3%). Colors correspond to clusters and are consistent throughout the figure.

```{r fig2a, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 2A: PROPORTION OF CD8⁺ T CELL SUBSETS
# ============================================================================

# Load pre-processed Seurat object
load("scRNA_harmony.RData")

cluster <- c("cluster2", "cluster3", "cluster11", "cluster19")
scales <- c(43, 39, 15, 3)
label <- paste(cluster, "(", scales, "%", ")", sep = "")
df <- data.frame(cluster = cluster, scales = scales, label = label)
Define color palette for CD8⁺ T cell clusters

cd8_cluster_colors <- c(
"#354D98", # cluster2 - deep blue
"#70A2C5", # cluster3 - light blue
"#F5B46B", # cluster11 - orange
"#DF5B3A" # cluster19 - red
)
Create 3D pie chart

pdf(file = here(output_dir, "Figure2A_pie_chart.pdf"), width = 10, height = 6)
pie3D(df$scales,
      radius = 1.5,
      labels = df$label,
      explode = 0.2,
      col = cd8_cluster_colors,
      main = "Proportion of CD8⁺ T Cell Subsets",
      labelcex = 1.2)
dev.off()
```

---

## **b. Feature expression in CD8⁺ T cell clusters**
  
**(b) Violin plots showing expression levels of 18 key marker genes across the four CD8⁺ T cell clusters.** Genes were selected based on differential expression analysis and include markers of T cell activation, exhaustion, and function. Cluster3 shows high expression of CCR7 and LEF1, while cluster19 expresses FCRL6 and LAG3.
```{r fig2b, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
============================================================================
FIGURE 2B: FEATURE EXPRESSION IN CD8⁺ T CELL CLUSTERS
============================================================================

Idents(scRNA_harmony) <- 'seurat_clusters'
CD8_T_cell <- subset(x = scRNA_harmony, idents = c("2", "3", "11", "19"))
Define marker genes

marker_genes <- c(
  "FCRL6", "ADGRG1", "LAG3", "LINC02446", "MAL", "PASK",
  "NELL2", "LEF1", "CCR7", "CD28", "MPZL3", "PTGER2",
  "TYMS", "PCLAF", "CD38", "MCM5", "CENPM", "RRM2"
)
Create violin plots

p2b <- VlnPlot(
  CD8_T_cell,
  features = marker_genes,
  pt.size = 0,
  ncol = 3,
  cols = cd8_cluster_colors
) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_line(color = "gray90", linewidth = 0.2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    strip.text = element_text(size = 11, face = "bold")
  ) +
  labs(
    title = "Marker Gene Expression in CD8⁺ T Cell Subsets",
    subtitle = "Violin plots showing expression of 18 key genes",
    y = "Expression Level"
  )

#Save high-resolution version
ggsave(
  filename = here(output_dir, "Figure2B_violin_plots.pdf"),
  plot = p2b,
  width = 12,
  height = 12,
  device = "pdf",
  dpi = 300
)
---

## **c. Dot plot of gene expression in cluster3**
  
**(c) Dot plot visualizing expression patterns of 45 selected genes in CD8⁺ T cell cluster3.** Color intensity represents average expression level, while dot size indicates the percentage of cells expressing each gene. The plot reveals distinct gene modules associated with T cell activation, cytotoxicity, and regulation.
```{r fig2c, fig.width=8, fig.height=10, warning=FALSE, message=FALSE}
============================================================================
FIGURE 2C: DOT PLOT OF GENE EXPRESSION IN CLUSTER3
============================================================================
  
#Define expanded gene set
dot_genes <- c(
  "FCRL6", "LINC01871", "C1orf21", "ADGRG1", "LAG3", "KLRG1", "LINC02446",
  "KLRC4", "ARMH1", "LYPD3", "MAL", "SCML1", "BTBD11", "DNAJB4", "MYC",
  "ANKRD37", "NELL2", "LEF1", "CCR7", "PASK", "GZMK", "LYAR", "IFNG-AS1",
  "FXYD2", "JUNB", "DUSP1", "TYMS", "PCLAF", "CD38", "MCM5", "CENPM",
  "RRM2", "TXNDC5", "PTTG1", "TK1", "NUSAP1", "CLSPN", "CENPU", "GINS2", "CDT1"
)

#Create dot plot
p2c <- DotPlot(
  CD8_T_cell,
  features = dot_genes,
  assay = 'RNA'
) +
  coord_flip() +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 9, face = "italic"),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 8),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  scale_color_gradientn(
    values = seq(0, 1, 0.2),
    colors = c('#330066', '#336699', '#66CC66', '#FFCC33'),
    name = "Average Expression"
  ) +
  scale_size_continuous(
    range = c(2, 8),
    name = "Percent Expressed"
  ) +
  labs(
    title = "Gene Expression Patterns in Cluster3",
    subtitle = "45 selected genes showing distinct expression modules"
  )

#Save high-resolution version
ggsave(
  filename = here(output_dir, "Figure2C_dot_plot.pdf"),
  plot = p2c,
  width = 8,
  height = 10,
  device = "pdf",
  dpi = 300
)
```

---
  
## **d. Differential gene expression in cluster3**
  
**(d) Volcano plot comparing gene expression between Sjögren's syndrome patients and healthy controls in CD8⁺ T cell cluster3.** Genes with adjusted p-value < 0.05 and |log2 fold change| > 0.15 are considered significant. Key genes GIMAP5, GIMAP4, and SNHG25 are highlighted, suggesting their potential role in SjS pathogenesis.
```{r fig2d, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
============================================================================
FIGURE 2D: DIFFERENTIAL GENE EXPRESSION IN CLUSTER3
============================================================================

#Subset cluster3
Idents(CD8_T_cell) <- 'seurat_clusters'
sub_cluster3 <- subset(CD8_T_cell, idents = "3")
Find differentially expressed genes

message("Identifying differentially expressed genes in cluster3...")
cluster3.markers <- FindMarkers(
sub_cluster3,
ident.1 = "SjS",
ident.2 = "HC",
group.by = "Group",
logfc.threshold = 0,
min.pct = 0.1
)

#Prepare data for volcano plot
cluster3.markers$names <- rownames(cluster3.markers)
cluster3.markersDifference<−cluster3.markerspct.1 - cluster3.markers$pct.2

#Classify genes
cluster3.markers$group <- "no"
cluster3.markersgroup[cluster3.markersavg_log2FC >= 0.25 &
cluster3.markers$Difference >= 0.1 &
cluster3.markers$pct.2 <= 0.5] <- "up"
cluster3.markersgroup[cluster3.markersavg_log2FC <= (-0.25) &
cluster3.markers$Difference <= -0.1 &
cluster3.markers$pct.1 <= 0.5] <- "down"

#Label key genes
label_genes <- c("GIMAP5", "GIMAP4", "SNHG25")
cluster3.markers$label <- ifelse(
cluster3.markers$names %in% label_genes,
cluster3.markers$names,
""
)

#Create volcano plot
p2d <- ggplot(cluster3.markers, aes(x = Difference, y = avg_log2FC)) +
geom_point(size = 1.5, aes(color = group, alpha = group)) +
scale_color_manual(
name = "Expression",
values = c("up" = "#cb5a54", "down" = "#1176bd", "no" = "#7b7f80"),
labels = c("Upregulated", "Downregulated", "Not significant")
) +
scale_alpha_manual(
values = c("up" = 0.8, "down" = 0.8, "no" = 0.3),
guide = "none"
) +
geom_label_repel(
aes(label = label),
fill = "#cb7935",
color = "white",
segment.size = 0.25,
size = 4,
max.overlaps = 20,
fontface = "bold",
label.size = NA,
box.padding = 0.5,
min.segment.length = 0
) +
geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.5) +
geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.5) +
theme_classic(base_size = 12) +
theme(
plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
axis.title = element_text(size = 12, face = "bold"),
axis.text = element_text(size = 10),
legend.position = "right",
legend.title = element_text(face = "bold"),
panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
) +
labs(
title = "Differential Gene Expression in Cluster3 (SjS vs HC)",
subtitle = "Volcano plot highlighting genes with significant expression changes",
x = "∆ Percentage Difference (pct.1 - pct.2)",
y = "Log₂ Fold Change"
) +
guides(color = guide_legend(override.aes = list(size = 3)))

#Save high-resolution version
ggsave(
filename = here(output_dir, "Figure2D_volcano_plot.pdf"),
plot = p2d,
width = 8,
height = 7,
device = "pdf",
dpi = 300
)
```

---

## **e. Differential gene expression in cluster19**
  
**(e) Volcano plot comparing gene expression between Sjögren's syndrome patients and healthy controls in CD8⁺ T cell cluster19.** Genes with adjusted p-value < 0.05 and |log2 fold change| > 0.15 are considered significant. Key genes GIMAP5, GIMAP4, and SNHG25 are highlighted, suggesting their potential role in SjS pathogenesis.
```{r fig2d, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
============================================================================
FIGURE 2D: DIFFERENTIAL GENE EXPRESSION IN CLUSTER3
============================================================================

#Subset cluster19
Idents(CD8_T_cell) <- 'seurat_clusters'
sub_cluster19 <- subset(CD8_T_cell, idents = "19")
Find differentially expressed genes

message("Identifying differentially expressed genes in cluster19...")
cluster19.markers <- FindMarkers(
sub_cluster19,
ident.1 = "SjS",
ident.2 = "HC",
group.by = "Group",
logfc.threshold = 0,
min.pct = 0.1
)

#Prepare data for volcano plot
cluster19.markers$names <- rownames(cluster19.markers)
cluster19.markersDifference<−cluster19.markerspct.1 - cluster19.markers$pct.2

#Classify genes
cluster19.markers$group <- "no"
cluster19.markersgroup[cluster19.markersavg_log2FC >= 0.25 &
cluster19.markers$Difference >= 0.1 &
cluster19.markers$pct.2 <= 0.5] <- "up"
cluster19.markersgroup[cluster19.markersavg_log2FC <= (-0.25) &
cluster19.markers$Difference <= -0.1 &
cluster19.markers$pct.1 <= 0.5] <- "down"

#Label key genes
label_genes <- c("GIMAP5", "GIMAP4", "SNHG25")
cluster19.markers$label <- ifelse(
cluster19.markers$names %in% label_genes,
cluster19.markers$names,
""
)

#Create volcano plot
p2d <- ggplot(cluster19.markers, aes(x = Difference, y = avg_log2FC)) +
geom_point(size = 1.5, aes(color = group, alpha = group)) +
scale_color_manual(
name = "Expression",
values = c("up" = "#cb5a54", "down" = "#1176bd", "no" = "#7b7f80"),
labels = c("Upregulated", "Downregulated", "Not significant")
) +
scale_alpha_manual(
values = c("up" = 0.8, "down" = 0.8, "no" = 0.3),
guide = "none"
) +
geom_label_repel(
aes(label = label),
fill = "#cb7935",
color = "white",
segment.size = 0.25,
size = 4,
max.overlaps = 20,
fontface = "bold",
label.size = NA,
box.padding = 0.5,
min.segment.length = 0
) +
geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.5) +
geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.5) +
theme_classic(base_size = 12) +
theme(
plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
axis.title = element_text(size = 12, face = "bold"),
axis.text = element_text(size = 10),
legend.position = "right",
legend.title = element_text(face = "bold"),
panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
) +
labs(
title = "Differential Gene Expression in Cluster3 (SjS vs HC)",
subtitle = "Volcano plot highlighting genes with significant expression changes",
x = "∆ Percentage Difference (pct.1 - pct.2)",
y = "Log₂ Fold Change"
) +
guides(color = guide_legend(override.aes = list(size = 3)))

#Save high-resolution version
ggsave(
filename = here(output_dir, "Figure2D_volcano_plot.pdf"),
plot = p2d,
width = 8,
height = 7,
device = "pdf",
dpi = 300
)
```

---

## **f. GO enrichment analysis of CD8⁺ T cell clusters**

**(f) Dot plot showing Gene Ontology (GO) enrichment analysis of differentially expressed genes from four CD8⁺ T cell clusters (cluster2, cluster3, cluster11, cluster19).** The analysis reveals enrichment of immune-related pathways, T cell activation, and cytokine signaling, with cluster-specific enrichments highlighting distinct functional states.
```{r fig2f, fig.width=12, fig.height=9, warning=FALSE, message=FALSE}
============================================================================
FIGURE 2F: GO ENRICHMENT ANALYSIS
============================================================================

# Load differential expression results
message("Loading differential expression results...")
marker_files <- c(
cluster2 = "cluster2.csv",
cluster3 = "cluster3.csv",
cluster11 = "cluster11.csv",
cluster19 = "cluster19.csv"
)

#Load and process marker files
marker_data <- list()
for (clust_name in names(marker_files)) {
file_path <- marker_files[clust_name]
if (file.exists(file_path)) {
df <- read.csv(file_path, header = TRUE)
colnames(df)[1] <- "SYMBOL"
# Filter significant genes
sig_genes <- df %>%
filter(p_val_adj < 0.01) %>%
filter(abs(avg_log2FC) > 0.25) %>%
pull(SYMBOL)
if (length(sig_genes) > 0) {
  # Convert to ENTREZID
  conv <- tryCatch({
    bitr(sig_genes,
         fromType = "SYMBOL",
         toType = "ENTREZID",
         OrgDb = org.Hs.eg.db)
  }, error = function(e) {
    NULL
  })
  
  if (!is.null(conv) && nrow(conv) > 0) {
    marker_data[[clust_name]] <- conv$ENTREZID
  }
}
}
}

#Run GO enrichment analysis
if (length(marker_data) > 0) {
message("Running GO enrichment analysis...")
go_results <- compareCluster(
geneClusters = marker_data,
fun = "enrichGO",
OrgDb = org.Hs.eg.db,
ont = "BP", # Biological Process
pvalueCutoff = 0.05,
pAdjustMethod = "BH",
qvalueCutoff = 0.2,
readable = TRUE
)

#Create dot plot
p2f <- dotplot(go_results,
showCategory = 8,
includeAll = FALSE,
font.size = 10) +
theme_minimal(base_size = 12) +
theme(
plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
axis.title = element_text(size = 12, face = "bold"),
axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
axis.text.y = element_text(size = 10),
legend.position = "right",
legend.title = element_text(face = "bold", size = 10),
legend.text = element_text(size = 8),
strip.text = element_text(size = 11, face = "bold"),
panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
panel.grid.minor = element_blank(),
panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
) +
labs(
title = "GO Enrichment Analysis of CD8⁺ T Cell Clusters",
subtitle = "Biological processes enriched in differentially expressed genes",
x = "Gene Ratio",
y = "GO Biological Process"
) +
scale_color_gradientn(
colors = c("blue", "green", "yellow", "red"),
limits = c(0, 0.05),
name = "Adjusted p-value"
) +
scale_size_continuous(
range = c(3, 8),
name = "Gene Count"
)

#Save high-resolution version
ggsave(
filename = here(output_dir, "Figure2F_GO_enrichment.pdf"),
plot = p2f,
width = 12,
height = 9,
device = "pdf",
dpi = 300
)
```

---

## **Session Information**

```{r session_info, echo=FALSE}
# ============================================================================
# SESSION INFORMATION
# ============================================================================
sessioninfo::session_info()
```
