---
title: "Figure 3. Functional characterization of gene signatures in Sjögren's syndrome"
author: "Cheng"
date: "2026/04/21"
output:
  html_document:
    toc: false
    toc_float: false
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
    fig_caption: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 1. INITIALIZATION AND SETUP
# ============================================================================
knitr::opts_chunk$set(
  echo = TRUE,           
  warning = TRUE,        
  message = TRUE,        
  error = TRUE,          
  fig.width = 10,        
  fig.height = 7,        
  fig.align = "center",  
  cache = FALSE,        
  dpi = 300              
)

# 清理工作空间
rm(list = ls())
gc()

# 设置输出目录
output_dir <- "."
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 设置随机种子保证可重复性
set.seed(123)
```

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 2. LOAD REQUIRED LIBRARIES
# ============================================================================
#Single-cell analysis
library(Seurat)
library(AUCell)
library(Nebulosa)

#Gene set analysis
library(GSEABase)
library(clusterProfiler)
library(GSVA)
library(GEOquery)

#Visualization
library(ggplot2)
library(ggpubr)
library(patchwork)
library(ggrepel)
library(viridis)
library(RColorBrewer)
library(cowplot)
library(ggsci)

#Data processing
library(dplyr)
library(tidyr)
library(tibble)
library(reshape2)
library(stringr)

#Functional packages
library(here)
library(hugene10sttranscriptcluster.db)
library(hgu133plus2.db)
library(org.Hs.eg.db)

#Check and install missing packages
required_packages <- c("Seurat", "AUCell", "GSVA", "GEOquery", "ggplot2", "dplyr")
missing_packages <- required_packages[!required_packages %in% installed.packages()]
if (length(missing_packages) > 0) {
message("Installing missing packages: ", paste(missing_packages, collapse = ", "))
BiocManager::install(missing_packages, update = FALSE)
}
```

<br>

# **Figure 3: Functional validation of Sjögren's syndrome gene signatures**

**Figure Legend:** Functional validation of Sjögren's syndrome-associated gene signatures identified in specific T cell subsets. (a-b) AUCell analysis showing activity scores of gene signatures in (a) CD8⁺ T cell cluster3 and (b) γδ T cell cluster15. (c) GSVA enrichment scores of cluster19 signature across six independent Sjögren's syndrome transcriptomic datasets, comparing patients (PSS) vs. healthy controls (HC).

---
## **a. AUCell analysis of CD8⁺ T cell cluster3 signature**
**(a) Activity scores of the cluster3-specific gene signature in CD8⁺ T cells.** AUCell calculates the activity of the gene signature in each cell based on the ranking of gene expression. UMAP visualization (left) shows spatial distribution of signature activity, while violin plots (right) compare activity scores across CD8⁺ T cell clusters.

```{r fig3a, fig.width=6, fig.height=5, warning=FALSE, message=FALSE}
============================================================================
FIGURE 3A: AUCELL ANALYSIS - CD8+ T CELL CLUSTER3
============================================================================

#Load necessary data
library(AUCell)
library(clusterProfiler)

library(ggplot2)
library(Seurat)
library(GSEABase)

# AUCell_buildRankings
# 计算gene signatures分值
cells_rankings <- AUCell_buildRankings(sub_CD8_T_cell@assays$RNA@counts, nCores=1, plotStats=TRUE)  

## 制作基因集
geneSets <- cluster3_a
extraGeneSets <- c(GeneSet(sample(cluster3_a),setName="CD8_gene"))

geneSets1 <- GeneSetCollection(extraGeneSets)
names(geneSets1) 

## 计算AUC
cells_AUC <- AUCell_calcAUC(geneSets1, cells_rankings)
cells_AUC

pdf('Figure3A_cluster3_cells_assignment.pdf',width = 6, height = 5)
cells_assignment <- AUCell_exploreThresholds(cells_AUC, plotHist=TRUE, assign=TRUE) 
dev.off()
```

---

```{r fig3b, fig.width=6, fig.height=5, warning=FALSE, message=FALSE}
============================================================================
FIGURE 3B: AUCELL ANALYSIS - CD8+ T CELL CLUSTER19
============================================================================

#Load necessary data
library(AUCell)
library(clusterProfiler)

library(ggplot2)
library(Seurat)
library(GSEABase)

# AUCell_buildRankings
# 计算gene signatures分值
cells_rankings <- AUCell_buildRankings(sub_CD8_T_cell@assays$RNA@counts, nCores=1, plotStats=TRUE)  

## 制作基因集
geneSets <- cluster19_a
extraGeneSets <- c(GeneSet(sample(cluster19_a),setName="CD8_gene"))

geneSets1 <- GeneSetCollection(extraGeneSets)
names(geneSets1) 

## 计算AUC
cells_AUC <- AUCell_calcAUC(geneSets1, cells_rankings)
cells_AUC

pdf('Figure3_cluster19_cells_assignment.pdf',width = 6, height = 5)
cells_assignment <- AUCell_exploreThresholds(cells_AUC, plotHist=TRUE, assign=TRUE) 
dev.off()
```

---

---

## **d. Multi-dataset validation of cluster19 signature**

**(d) GSVA enrichment scores of the cluster19 gene signature across six independent Sjögren's syndrome transcriptomic datasets.** Validation in bulk RNA-seq datasets (GSE127952, GSE159574, GSE208260, GSE97614, GSE7451, GSE40611) shows consistent upregulation of the signature in Sjögren's syndrome patients compared to healthy controls.


```{r fig3d, fig.width=6, fig.height=5, warning=FALSE, message=FALSE}
============================================================================
FIGURE 3A: GSVA ANALYSIS ACROSS MULTIPLE DATASETS - CD8+ T CELL CLUSTER19
============================================================================

#Load necessary data
library(GEOquery)
library(GSVA)
library(ggplot2)
library(ggpubr)

gs <- read.table("/data/shumin/2022-08-29-PSS-Project/singlecell-score/cluster19_gs.txt", header = T)
geneGSVA <- as.data.frame(gs$Gene)
colnames(geneGSVA) <- 'gs'



