---
title: "Figure 1. Single-cell landscape and annotation of peripheral blood immune cells in Sjögren's syndrome"
author: "Your Name"
date: "2026/04/21"
output: 
  html_document:
    toc: false
    toc_float: false
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
    fig_caption: true
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 1. INITIALIZATION AND SETUP
# ============================================================================
knitr::opts_chunk$set(
  echo = TRUE,           # 显示代码
  warning = TRUE,        # 显示警告
  message = TRUE,        # 显示消息
  error = TRUE,          # 显示错误
  fig.width = 10,        # 默认图形宽度
  fig.height = 7,        # 默认图形高度
  fig.align = "center",  # 图形居中
  cache = FALSE,         # 不缓存结果（首次运行可设为TRUE加速）
  dpi = 300              # 图形分辨率
)

# 清理工作空间
rm(list = ls())
gc()

# 设置输出目录
output_dir <- "./output/"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# 设置随机种子保证可重复性
set.seed(2024)
```

```{r libraries, include=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# 2. LOAD REQUIRED LIBRARIES
# ============================================================================
# 核心分析包
library(Seurat)          # 单细胞分析
library(SingleR)         # 细胞注释
library(Nebulosa)        # 基因表达密度图

# 可视化包
library(ggplot2)         # 基础绘图
library(patchwork)       # 图形组合
library(RColorBrewer)    # 颜色调色板
library(viridis)         # 科学配色
library(cowplot)         # 图形修饰

# 数据处理包
library(dplyr)           # 数据操作
library(tidyr)           # 数据整理
library(reshape2)        # 数据重塑
library(tibble)          # 现代数据框

# 功能包
library(here)            # 路径管理
library(knitr)           # 动态文档
library(kableExtra)      # 表格美化

# 检查并安装缺失包
required_packages <- c("Seurat", "SingleR", "ggplot2", "dplyr", "patchwork")
missing_packages <- required_packages[!required_packages %in% installed.packages()]
if (length(missing_packages) > 0) {
  message("Installing missing packages: ", paste(missing_packages, collapse = ", "))
  BiocManager::install(missing_packages, update = FALSE)
}
```

<br>

# **Figure 1: Single-cell atlas of PBMCs in Sjögren's syndrome**

**Figure Legend:** Single-cell RNA sequencing analysis of peripheral blood mononuclear cells (PBMCs) from 5 healthy controls (HC) and 5 Sjögren's syndrome (SjS) patients. (a) t-SNE visualization of 57,626 cells colored by 22 unsupervised clusters. (b) Major cell type annotation using SingleR reference mapping. (c) Joint density plots of key lineage-specific marker genes. (d) Heatmap showing expression patterns of Sjögren's syndrome-associated gene modules across 22 refined cell subtypes.

---

## **a. t-SNE clustering of all PBMCs**

**(a) t-SNE visualization of 57,626 peripheral blood mononuclear cells (PBMCs) from healthy controls (HC) and Sjögren's syndrome (SjS) patients.** Cells are colored by 22 distinct clusters identified through unsupervised Louvain clustering. The analysis reveals comprehensive immune cell heterogeneity in both health and disease states.

```{r fig1a, fig.width=9, fig.height=8, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 1A: t-SNE CLUSTERING
# ============================================================================

# Load pre-processed Seurat object
message("Loading Seurat object...")
if (!file.exists("scRNA_harmony.RData")) {
  stop("File 'scRNA_harmony.RData' not found. Please ensure it's in the working directory.")
}
load("scRNA_harmony.RData")

# Validate object structure
message("Validating Seurat object...")
print(paste("Number of cells:", ncol(scRNA_harmony)))
print(paste("Number of features:", nrow(scRNA_harmony)))
print(paste("Available reductions:", names(scRNA_harmony@reductions)))

# Define color palette for 22 clusters
cluster_colors <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B",
  "#E377C2", "#7F7F7F", "#BCBD22", "#17BECF", "#AEC7E8", "#FFBB78",
  "#98DF8A", "#FF9896", "#C5B0D5", "#C49C94", "#F7B6D2", "#C7C7C7",
  "#DBDB8D", "#9EDAE5", "#393B79", "#5254A3"
)

# Create t-SNE plot
p1a <- DimPlot(
  scRNA_harmony,
  reduction = "tsne",
  group.by = "seurat_clusters",
  label = TRUE,
  label.size = 4,
  pt.size = 0.3,
  cols = cluster_colors,
  repel = TRUE
) +
  labs(
    title = "Unsupervised clustering of 57,626 PBMCs",
    subtitle = "22 distinct clusters identified by Louvain algorithm",
    x = "t-SNE 1",
    y = "t-SNE 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_line(color = "gray90", linewidth = 0.2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  guides(color = guide_legend(
    title = "Cluster ID",
    ncol = 2,
    override.aes = list(size = 3)
  ))

# Save high-resolution version
ggsave(
  filename = here(output_dir, "Figure1A_tSNE_clusters.pdf"),
  plot = p1a,
  width = 9,
  height = 8,
  device = "pdf",
  dpi = 300
)

# Save PNG for quick viewing
ggsave(
  filename = here(output_dir, "Figure1A_tSNE_clusters.png"),
  plot = p1a,
  width = 9,
  height = 8,
  dpi = 300,
  bg = "white"
)

# Display plot
p1a

# Print cluster statistics
cluster_stats <- scRNA_harmony@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    n_cells = n(),
    percent = round(n() / ncol(scRNA_harmony) * 100, 2)
  ) %>%
  arrange(seurat_clusters)

message("\nCluster composition:")
print(cluster_stats)
```

**Table: Cluster size distribution**
```{r cluster_table, echo=FALSE}
kable(cluster_stats, caption = "Number of cells per cluster") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## **b. Major cell type annotation**

**(b) Cell type annotation at the major lineage level.** Annotation was performed using SingleR with the Human Primary Cell Atlas as reference. Colors represent 10 major immune cell lineages identified in PBMCs.

```{r fig1b, fig.width=9, fig.height=8, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 1B: MAJOR CELL TYPE ANNOTATION
# ============================================================================

# Load reference data
message("Loading reference data for SingleR annotation...")
if (!file.exists("HumanPrimaryCellAtlasData.Rdata")) {
  stop("Reference data file 'HumanPrimaryCellAtlasData.Rdata' not found.")
}
load("HumanPrimaryCellAtlasData.Rdata")
ref_data <- ref  # Rename for clarity

# Prepare data for SingleR
message("Preparing data for SingleR annotation...")
sce_data <- GetAssayData(scRNA_harmony, layer = "data")
cluster_ids <- scRNA_harmony@meta.data$seurat_clusters

# Run SingleR annotation
message("Running SingleR annotation...")
singler_results <- SingleR(
  test = sce_data,
  ref = ref_data,
  labels = ref_data$label.main,
  method = "cluster",
  clusters = cluster_ids,
  assay.type.test = "logcounts",
  assay.type.ref = "logcounts"
)

# Save annotation results
annotation_df <- data.frame(
  Cluster = rownames(singler_results),
  Predicted_Label = singler_results$labels,
  Score = singler_results$scores,
  stringsAsFactors = FALSE
)

write.csv(
  annotation_df,
  file = here(output_dir, "SingleR_annotation_results.csv"),
  row.names = FALSE
)

message("SingleR annotation completed. Results saved to CSV.")

# Define manual annotation based on SingleR results
# This mapping combines SingleR predictions with domain knowledge
major_type_mapping <- c(
  "0" = "CD4+ T cell",      # Helper T cells
  "1" = "CD4+ T cell",      # Regulatory T cells
  "2" = "CD8+ T cell",      # Cytotoxic T cells
  "3" = "CD8+ T cell",      # Memory CD8+ T cells
  "4" = "NK Cell",          # Natural Killer cells
  "5" = "NK Cell",          # Activated NK cells
  "6" = "Monocytes",        # Classical monocytes
  "7" = "Monocytes",        # Non-classical monocytes
  "8" = "B Cell",           # B lymphocytes
  "9" = "γδ T cell",        # Gamma delta T cells
  "10" = "γδ T cell",       # TRGC2+ γδ T cells
  "11" = "CD8+ T cell",     # Effector memory CD8+ T cells
  "12" = "CD4+ T cell",     # Naive CD4+ T cells
  "13" = "Macrophages",     # Macrophages
  "14" = "B Cell",          # Regulatory B cells
  "15" = "γδ T cell",       # TRDC+ γδ T cells
  "16" = "Platelets",       # Platelets
  "17" = "CD4+ T cell",     # Helios+ Tregs
  "18" = "DC cell",         # Dendritic cells
  "19" = "CD8+ T cell",     # Exhausted CD8+ T cells
  "20" = "Monocytes",       # Inflammatory monocytes
  "21" = "CMP"              # Common myeloid progenitors
)

# Add annotations to Seurat object
scRNA_harmony@meta.data$Major.CellType <- factor(
  major_type_mapping[as.character(scRNA_harmony@meta.data$seurat_clusters)],
  levels = c(
    "CD4+ T cell", "CD8+ T cell", "γδ T cell", "NK Cell",
    "B Cell", "Monocytes", "Macrophages", "DC cell",
    "Platelets", "CMP"
  )
)

# Define color scheme for major cell types
major_type_colors <- c(
  "CD4+ T cell" = "#E41A1C",    # Red
  "CD8+ T cell" = "#377EB8",    # Blue
  "γδ T cell" = "#4DAF4A",      # Green
  "NK Cell" = "#984EA3",        # Purple
  "B Cell" = "#FF7F00",         # Orange
  "Monocytes" = "#FFFF33",      # Yellow
  "Macrophages" = "#A65628",    # Brown
  "DC cell" = "#F781BF",        # Pink
  "Platelets" = "#999999",      # Gray
  "CMP" = "#66C2A5"             # Teal
)

# Create annotation plot
p1b <- DimPlot(
  scRNA_harmony,
  reduction = "tsne",
  group.by = "Major.CellType",
  label = FALSE,
  pt.size = 0.3,
  cols = major_type_colors,
  order = TRUE
) +
  labs(
    title = "Major cell type annotation",
    subtitle = "10 major immune lineages identified by SingleR",
    x = "t-SNE 1",
    y = "t-SNE 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_line(color = "gray90", linewidth = 0.2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  guides(color = guide_legend(
    title = "Major Cell Type",
    ncol = 1,
    override.aes = list(size = 4, alpha = 1)
  ))

# Save plots
ggsave(
  filename = here(output_dir, "Figure1B_major_types.pdf"),
  plot = p1b,
  width = 9,
  height = 8,
  device = "pdf"
)

# Display plot
p1b

# Print annotation summary
annotation_summary <- scRNA_harmony@meta.data %>%
  group_by(Major.CellType) %>%
  summarise(
    n_cells = n(),
    percent = round(n() / ncol(scRNA_harmony) * 100, 2),
    n_clusters = n_distinct(seurat_clusters)
  ) %>%
  arrange(desc(n_cells))

message("\nMajor cell type composition:")
print(annotation_summary)
```

**Table: Major cell type distribution**
```{r major_type_table, echo=FALSE}
kable(annotation_summary, caption = "Distribution of major cell types") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## **c. Marker gene expression validation**

**(c) Joint density plots of key lineage-specific marker genes.** Each panel shows co-expression patterns of canonical markers validating cell type identities. Density visualization highlights both expression level and spatial distribution in t-SNE space.

```{r fig1c, fig.width=14, fig.height=12, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 1C: MARKER GENE EXPRESSION DENSITY
# ============================================================================

# Define marker gene sets for each major lineage
marker_sets <- list(
  "Monocyte/Macrophage" = list(
    genes = c("FCN1", "S100A12", "CD68", "CD163", "MS4A7"),
    title = "Monocyte/Macrophage\n(FCN1, S100A12, CD68, CD163, MS4A7)"
  ),
  "Dendritic Cells" = list(
    genes = c("CD1C", "FCER1A", "CLEC10A"),
    title = "Dendritic Cells\n(CD1C, FCER1A, CLEC10A)"
  ),
  "NK Cells" = list(
    genes = c("NKG7", "CD247", "KLRC1", "KLRF1"),
    title = "Natural Killer Cells\n(NKG7, CD247, KLRC1, KLRF1)"
  ),
  "Platelets" = list(
    genes = c("ITGA2B", "PF4", "GP1BA", "TUBB1"),
    title = "Platelets\n(ITGA2B, PF4, GP1BA, TUBB1)"
  ),
  "B Cells" = list(
    genes = c("MS4A1", "CD79A", "CD79B"),
    title = "B Lymphocytes\n(MS4A1, CD79A, CD79B)"
  ),
  "T Cells" = list(
    genes = c("CD3D", "CD3E"),
    title = "T Lymphocytes\n(CD3D, CD3E)"
  ),
  "CD4+ T Cells" = list(
    genes = c("CD3E", "CD4"),
    title = "CD4+ T Cells\n(CD3E, CD4)"
  ),
  "CD8+ T Cells" = list(
    genes = c("CD8A", "CD8B"),
    title = "CD8+ T Cells\n(CD8A, CD8B)"
  ),
  "γδ T Cells" = list(
    genes = c("TRDC", "TRGC1", "TRGC2"),
    title = "γδ T Cells\n(TRDC, TRGC1, TRGC2)"
  )
)

# Create individual density plots
create_density_plot <- function(seurat_obj, gene_set, plot_title) {
  p <- plot_density(
    seurat_obj,
    gene_set,
    joint = TRUE,
    reduction = "tsne",
    size = 0.5
  )[[length(gene_set)]] +  # Get the joint density plot
    
    labs(title = plot_title) +
    
    # Customize appearance
    theme_minimal() +
    theme(
      plot.title = element_text(
        size = 12,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 5)
      ),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none",
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_rect(
        fill = NA,
        color = "black",
        linewidth = 0.8
      ),
      plot.margin = margin(5, 5, 5, 5)
    ) +
    
    # Use viridis color scale
    scale_color_viridis_c(option = "plasma", direction = -1) +
    
    # Remove axis labels
    xlab(NULL) +
    ylab(NULL)
  
  return(p)
}

# Generate all density plots
density_plots <- list()
for (i in seq_along(marker_sets)) {
  cell_type <- names(marker_sets)[i]
  genes <- marker_sets[[i]]$genes
  title <- marker_sets[[i]]$title
  
  message(sprintf("Creating density plot for %s...", cell_type))
  
  density_plots[[cell_type]] <- create_density_plot(
    scRNA_harmony,
    genes,
    title
  )
}

# Arrange plots in a 3x3 grid
p1c <- wrap_plots(density_plots, ncol = 3) +
  plot_annotation(
    title = "Lineage-specific marker gene expression",
    subtitle = "Joint density plots validate cell type identities",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

# Save combined plot
ggsave(
  filename = here(output_dir, "Figure1C_marker_density.pdf"),
  plot = p1c,
  width = 14,
  height = 12,
  device = "pdf"
)

# Display the plot
p1c

# Save individual plots for potential supplemental figures
for (cell_type in names(marker_sets)) {
  ggsave(
    filename = here(output_dir, sprintf("marker_%s.pdf", gsub("/", "_", cell_type))),
    plot = density_plots[[cell_type]],
    width = 4,
    height = 4
  )
}

message("All marker density plots created and saved.")
```

---

## **d. Sjögren's syndrome gene module analysis**

**(d) Heatmap showing z-score normalized expression of 15 Sjögren's syndrome-associated gene modules across 22 refined cell subtypes.** Columns are ordered by major lineage, rows by gene module membership. Dashed lines separate different cell lineages and gene modules.

```{r fig1d, fig.width=12, fig.height=10, warning=FALSE, message=FALSE}
# ============================================================================
# FIGURE 1D: PSS GENE MODULE HEATMAP
# ============================================================================

# Define refined cell subtypes
subtype_mapping <- c(
  "0" = "Naive CD4+ T cell",
  "1" = "CD25hi Foxp3hi CD4+ Treg",
  "2" = "ZNF683+ CD8+ TF",
  "3" = "Naive CD8+ T cell",
  "4" = "KLRF1hi NK cell",
  "5" = "KLRC1lo NK cell",
  "6" = "CD14+ Monocytes",
  "7" = "S100A12+ Monocytes",
  "8" = "GC B cell",
  "9" = "TRGC2hi γδ T cell",
  "10" = "TRGC2me γδ T cell",
  "11" = "GZMKhi CD8+ TEM",
  "12" = "Foxp3hi CD4+ Treg",
  "13" = "Macrophage",
  "14" = "CD19+ CD24hi CD27+ Bregs",
  "15" = "TRDC γδ T cell",
  "16" = "Platelets",
  "17" = "Helios Foxp3lo CD4+ Treg",
  "18" = "DC cell",
  "19" = "CTLA-4 CD8+ Inhibitory",
  "20" = "GZMB+ Monocytes",
  "21" = "CMP"
)

# Add refined subtypes to metadata
scRNA_harmony@meta.data$Minor.CellType <- factor(
  subtype_mapping[as.character(scRNA_harmony@meta.data$seurat_clusters)]
)

# Define sample groups (HC vs SjS)
sample_group_mapping <- c(
  "sample1" = "HC", "sample2" = "HC", "sample3" = "HC",
  "sample4" = "HC", "sample5" = "HC", "sample6" = "SjS",
  "sample7" = "SjS", "sample8" = "SjS", "sample9" = "SjS",
  "sample10" = "SjS"
)

scRNA_harmony@meta.data$Group <- factor(
  sample_group_mapping[as.character(scRNA_harmony@meta.data$orig.ident)],
  levels = c("HC", "SjS")
)

# Load PSS gene modules
message("Loading PSS gene modules...")
if (!file.exists("PSS_gene.txt")) {
  stop("Gene module file 'PSS_gene.txt' not found.")
}

pss_genes <- read.table(
  "PSS_gene.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = "",
  comment.char = ""
)

# Remove duplicates
pss_genes <- pss_genes[!duplicated(pss_genes$gene), ]

# Ensure proper factor levels
pss_genes$gene <- factor(pss_genes$gene, levels = unique(pss_genes$gene))
pss_genes$group <- factor(
  pss_genes$group,
  levels = paste0("g", 1:15)
)

# Calculate average expression per subtype
Idents(scRNA_harmony) <- "Minor.CellType"

avg_expression <- AverageExpression(
  scRNA_harmony,
  assays = "RNA",
  features = levels(pss_genes$gene),
  group.by = "Minor.CellType",
  slot = "data"
)$RNA

# Order cell types by major lineage
celltype_info <- unique(scRNA_harmony@meta.data[, c("Major.CellType", "Minor.CellType")])
celltype_info$Major.CellType <- factor(
  celltype_info$Major.CellType,
  levels = c(
    "Macrophages", "Monocytes", "DC cell", "NK Cell",
    "Platelets", "B Cell", "CD4+ T cell", "CD8+ T cell",
    "γδ T cell", "CMP"
  )
)

celltype_info <- celltype_info[order(celltype_info$Major.CellType), ]
celltype_info$Minor.CellType <- factor(
  celltype_info$Minor.CellType,
  levels = unique(celltype_info$Minor.CellType)
)

# Reorder expression matrix
avg_expression <- avg_expression[, levels(celltype_info$Minor.CellType)]

# Z-score normalization (by gene)
zscore_matrix <- t(scale(t(avg_expression)))

# Clip extreme values
zscore_matrix[zscore_matrix > 3] <- 3
zscore_matrix[zscore_matrix < -2.5] <- -2.5

# Prepare data for plotting
plot_data <- zscore_matrix %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  mutate(Gene = factor(Gene, levels = rev(levels(pss_genes$gene)))) %>%
  pivot_longer(
    cols = -Gene,
    names_to = "CellType",
    values_to = "Zscore"
  ) %>%
  mutate(
    CellType = factor(CellType, levels = levels(celltype_info$Minor.CellType))
  )

# Create color palette
heatmap_palette <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)

# Create main heatmap
heatmap_plot <- ggplot(plot_data, aes(x = as.numeric(CellType), y = Gene, fill = Zscore)) +
  geom_tile(color = NA, linewidth = 0) +
  
  # Color scale
  scale_fill_gradientn(
    colors = heatmap_palette,
    limits = c(-2.5, 3),
    na.value = "white",
    name = "Z-score"
  ) +
  
  # X-axis labels
  scale_x_continuous(
    expand = c(0, 0),
    breaks = seq(1, ncol(avg_expression), by = 2),
    labels = levels(celltype_info$Minor.CellType)[seq(1, ncol(avg_expression), by = 2)],
    sec.axis = dup_axis(
      breaks = seq(2, ncol(avg_expression), by = 2),
      labels = levels(celltype_info$Minor.CellType)[seq(2, ncol(avg_expression), by = 2)]
    )
  ) +
  
  scale_y_discrete(expand = c(0, 0)) +
  
  # Add separation lines
  geom_vline(
    xintercept = cumsum(table(celltype_info$Major.CellType)) + 0.5,
    linetype = "dashed",
    color = "gray40",
    linewidth = 0.5
  ) +
  
  geom_hline(
    yintercept = cumsum(rev(table(pss_genes$group)[-1])) + 0.5,
    linetype = "dashed",
    color = "gray40",
    linewidth = 0.5
  ) +
  
  # Theme customization
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x.top = element_text(
      angle = 90,
      hjust = 0,
      vjust = 0.5,
      size = 9
    ),
    axis.text.x.bottom = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5,
      size = 9
    ),
    axis.ticks = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

# Create gene annotation bar
gene_annotation <- pss_genes %>%
  mutate(
    x = 1,
    y = seq_len(n()),
    group = factor(group, levels = paste0("g", 1:15))
  )

# Define colors for gene modules
module_colors <- colorRampPalette(brewer.pal(12, "Set3"))(15)

annotation_plot <- ggplot(gene_annotation, aes(x = x, y = y, fill = group)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = gene),
    size = 2.8,
    family = "serif",
    fontface = "italic",
    hjust = 0,
    nudge_x = 0.5
  ) +
  scale_fill_manual(values = module_colors) +
  scale_x_continuous(expand = c(0, 2)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

# Combine plots
combined_plot <- annotation_plot + heatmap_plot +
  plot_layout(widths = c(1, 3)) +
  plot_annotation(
    title = "Sjögren's syndrome-associated gene module expression",
    subtitle = "Z-score normalized expression across 22 immune cell subtypes",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

# Save combined plot
ggsave(
  filename = here(output_dir, "Figure1D_PSS_heatmap.pdf"),
  plot = combined_plot,
  width = 12,
  height = 10,
  device = "pdf"
)

# Display plot
combined_plot

# Save expression data
write.csv(
  zscore_matrix,
  file = here(output_dir, "PSS_gene_expression_matrix.csv"),
  row.names = TRUE
)

message("PSS gene module heatmap created and data saved.")
```

---

## **Session Information**

```{r session_info, echo=FALSE}
# ============================================================================
# SESSION INFORMATION
# ============================================================================
sessioninfo::session_info()
```